### 2019-02-27 2-Chapters/4-Chapter/CrownCom_17.git/IoT_slotted.tex
Finishing Chapter:4

As explained before, setting up the future Internet of Things (IoT) networks will require to support more and more communicating devices.
In this Section, we prove that intelligent devices in unlicensed bands can use Multi-Armed Bandit (MAB) learning algorithms to improve resource exploitation.
%
We evaluate the performance of two classical MAB learning algorithms, \UCB{} and Thomson Sampling, to handle the decentralized decision-making of Spectrum Access, applied to IoT networks; as well as learning performance with a growing number of intelligent end-devices.
%
We show that using learning algorithms does help to fit more devices in such networks, even when all end-devices are intelligent and are dynamically changing channel.
In the studied scenario, stochastic MAB learning provides a up to $16\%$ gain in term of successful transmission probabilities, and has near optimal performance even in non-stationary and non-\iid{} settings with a majority of intelligent devices.

@@

The aim of this Section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases, and the stochastic hypothesis is more and more questionable.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could have been generated by devices using other standards as well.
We first evaluate the probability of collision if dynamic devices randomly select channels (naive approach), and if a centralized controller optimally distribute them in channels (ideal approach).
Then, these reference scenarios allow to evaluate the performance of \UCB{} and TS algorithms in a decentralized network, in terms of successful communication rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic.

\paragraph{Outline}
The rest of this Section is organized as follows. The system model is introduced in Section~\ref{sub:41:systemModel}. Reference policies are described in \ref{sub:41:threeReferencePolicies}, and sequential learning policies are introduced in \ref{sub:41:sequentialPolicies}.
Numerical results are presented in \ref{sub:41:numericalResults}.

@@

We present our system model, which consists of one gateway and many IoT devices, using a slotted protocol.

\paragraph{One gateway and many devices}
%
We consider the system model presented in Figure~\ref{fig:41:system_model1}, where a set of object sends uplink packets to the network gateway.
Messages can be sent for instance in the $433.5\;\mathrm{MHz}$ ISM band (as it is chosen for the implementation presented in Section~\ref{sec:4:gnuradio}).
The communication between IoT devices and this gateway is done through a simple pure ALOHA-based protocol where devices transmit uplink packets of fixed duration whenever they want.
%
The devices can transmit their packets in $K\geq 1$ channels (\emph{e.g.}, $K=4$). In the case where the gateway receives an uplink in one channel, it transmits an acknowledgement to the end-device in the same channel, after a fixed delay (of $1\;\mathrm{s}$ in our implementation).

These communications operate in unlicensed ISM bands and, consequently, suffer from interference generated by uncoordinated neighboring networks. This interfering traffic is uncontrolled, and can be unevenly distributed over the $K$ different channels.
%
To simulate networks designed for the Internet of Things (IoT), we consider a protocol with no sensing, no repetition of uplink messages, and where the gateway is in charge of sending back an acknowledgement, after some fixed-time delay (typically $1\;\mathrm{s}$), to any object who succeeded in sending successfully an uplink packet.
%
By considering a small number of wireless channels (\emph{e.g.}, $K=8$) and one PHY layer configuration (\emph{i.e.}, modulation, waveform, etc), and in case of a non-uniform traffic in the different channels,
the object can improve their usage of the network if they are able to \emph{learn} on the fly the best channels to use (\emph{i.e.}, the most vacant one).

@@

\paragraph{Slotted protocol}
%
For simplicity, and for consistency with the rest of this thesis, we only consider a discrete time protocol.
As illustrated in Figure \ref{fig:41:protocol}, we suppose a slotted protocol, in both time and frequency.
All devices share a synchronized time, and know in advance $K$, the finite number of available RF channels.
In each time slot, the devices try to send packets to the unique Base Station, which listens continuously to all channels, following an ALOHA-based communication (\ie, no sensing).
Each time slot is divided in two parts: first for uplink communications in which data packets are sent by end-devices to the base station. If only one packet is sent in this part of the slot, the base station can decode it and sends an acknowledgement to the device in the second part (on the same channel).
If two or more devices send an uplink packet in the same slot, the uplink packets collide (\ie, there is a \emph{collision}), and the acknowledgement \emph{Ack} is not transmitted.
This way, no collision can occur on the downlink messages, easing the analysis of collisions.

\begin{figure}[!t]
    \centering
    \includegraphics[scale=0.40]{protocol.eps}
    \caption{The considered time-frequency slotted protocol. Each frame is composed by a fix-duration \textcolor{blue}{uplink slot} in which the end-devices transmit their \textcolor{blue}{(uplink) packets}. If a packet is well received, the base station replies by transmitting an \textcolor{darkgreen}{\emph{Ack}}, after the ack delay.}
    \label{fig:41:protocol}
\end{figure}

\paragraph{Static vs dynamic devices}
%
There are two types of end-devices in the network:
\begin{itemize}
    \item
    \emph{Static} end-devices have poor RF abilities, and each of them uses only one channel to communicate with the base station. Their choice is assumed to be fixed in time (stationary) and independent (\iid). The traffic generated by these devices is considered as an interfering traffic for other devices.
    \item
    \emph{Dynamic} (or \emph{smart}) end-devices have richer RF abilities, they can use all the available channels, by quickly reconfiguring their RF transceiver on the fly. They can also store communication successes or failures they experienced in each channel, in order to change channel, possibly at every time slot.
\end{itemize}

There are $K \geq 1$ channels, $D \geq 0$ dynamic end-devices, and $S \geq 0$ static devices.
Furthermore, in channel $i \in \llbracket 1; K \rrbracket$ there are $0 \leq S_i \leq S$ static devices (so $S = \sum_{i=1}^{K} S_i$).
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (about $1000$ to $10000$, while $K$ is about $4$ to $50$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the base station, we assume devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for IoT networks\footnote{~Here, $K$ will typically be about $20$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.

@@

\paragraph{MAB model}

We consider the network from the point of view of one dynamic device. Every times the dynamic device has to communicate with the gateway,
it has to choose one channel (at each transmission $t \geq 1, t \in \mathbb{N}$), denoted as $C(t) = k \in\{1,\dots,K\}$.
Then, the dynamic device starts waiting in this channel $C(t)$ for an acknowledgement sent by the gateway.
Before sending another message (\emph{i.e.}, at time $t+1$), the dynamic device knows if it received or not this \emph{ACK} message.
%
For this reason, selecting channel (or arm) $k$ at time $t$ yields a (random) feedback, called a \emph{reward}, $r_k(t) \in \{0,1\}$, being $0$ if no \emph{ACK} was received before the next message, or $1$ if \emph{ACK} was successfully received.
The goal of the dynamic device is to minimize its packet loss ratio, or equivalently, it is to maximize its cumulative reward,
$r_{1 \dots T} := \sum_{t = 1}^T r_{C(t)}(t),$
as it is usually done in MAB problems.

This problem is a special case of the stochastic MAB (see Chapter~\ref{chapter:2}).
% , where the sequence of rewards drawn from a given arm $k$ is assumed to be  \emph{i.i.d.}, under some distribution $\nu_k$, that has a mean $\mu_k$.
% Several types of reward distributions have been considered in the literature, and rewards are binary in our model.
We consider only Bernoulli distributions, in which $r_k(t) \sim \mathrm{Bern}(\mu_k)$, that is, $r_k(t) \in \{0,1\}$ and $\mathbb{P}(r_k(t) = 1) = \mu_k \in [0,1]$.
Contrary to many previous work done in the CR field (\emph{e.g.}, Opportunistic Spectrum Access),
the reward $r_k(t)$ does \emph{not} come from a sensing phase before sending the $t$-th message, as it would do for any ``listen-before-talk'' model.
Rewards come from receiving an acknowledgement from the gateway, between the $t$-th and $t+1$-th messages.

The problem parameters $\mu_1,\dots,\mu_K$ are of course unknown to each dynamic device, so to maximize its cumulated reward, it must learn the distributions of the channels, in order to be able to progressively focus on the best arm (\emph{i.e.}, the arm with largest mean).
%
This requires to tackle the so-called \emph{exploration-exploitation dilemma}: a device has to try all arms a sufficient number of times to get a robust estimate of their qualities, while not selecting the worst arms too many times.
%
The goal is thus to design a new or chose an existing simple sequential algorithm, to be applied identically by each dynamic device, in a fully distributed setting (each device runs its own algorithm, from its observations), in order to minimize collisions and maximize the fraction of successful transmissions of all the dynamic devices.

Before explaining how this goal presents similarity with a \emph{multi-armed bandit problem}, we present some natural baseline policies (\ie, algorithms).

@@

We present three different policies that will be used to assess the efficiency of the learning algorithms presented in the next Subsection.
The first one is naive but can be used in practice, while the two others are very efficient but require full knowledge on the system (\ie, an oracle) and are thus unpractical.

@@

We derive here the probability of having a successful transmission, for a dynamic device, in the case where all the dynamic devices make a purely random channel selection (\ie, uniform on $i \in \llbracket 1; K \rrbracket = \{1, \dots, K\}$).
This reflects a very naive policy that could be implemented by all the dynamic devices, and it provides a reference scenario to compare against.

In this case, for one dynamic device, a successful transmission happens if it is the only device to choose channel $i$, at that time slot.
The probability of successful transmission is computed as follows, because the $S_i$ static devices in each channel $i$ are assumed to be independent, and static and dynamic devices are assumed to \emph{not} transmit at each time $t$ with a fixed probability $1-p$:
\begin{equation}
    \Pr(\text{success}|\text{sent}) = \sum_{i=1}^{K} \underbrace{\Pr(\text{success}|\text{sent in channel}\;i)}_{\text{No one else sent in channel}\; i} \; \underbrace{\Pr(\text{sent in channel}\,i)}_{= 1/K, \text{by uniform choice}}
\end{equation}

All dynamic devices follow the same policy in this case, so the probability of transmitting at that time in channel $i$ for any dynamic device is $p / K$, and there are $D-1$ other dynamic devices.
As they are independent, the probability that no other dynamic device sent in $i$
% $\Pr(\text{no other dynamic device sent in}\;i)$,
is $q = \Pr(\bigcap_{k=1}^{D-1} \text{device}\;k\;\text{did not sent in}\;i) = \prod_{k=1}^{D-1} \Pr(\text{device}\;k\;\text{did not sent in}\;i)$. And $\Pr(\text{device}\;k\;\text{sent in}\;i) = p \times 1 / K$, by uniform choice on channels and the Bernoulli emission hypothesis. So $q = \prod_{k=1}^{D-1} (1 - p/K) = (1-p/K)^{D-1}$. Thus we can conclude,
%
\begin{align}\label{eq:41:strategynaive}
    \Pr(\text{success}|\text{sent})
    & = \sum_{i=1}^{K} \underbrace{(1 - p / K)^{D-1}}_{\text{No other dynamic device}} \times \underbrace{(1-p)^{S_i}}_{\text{No static device}} \times\; \frac{1}{K} \nonumber \\
    & = \frac{1}{K} \left(1-\frac{p}{K}\right)^{D-1} \sum_{i=1}^{K} (1-p)^{S_i} .
\end{align}
This expression \eqref{eq:41:strategynaive} is constant (in time), and easy to compute numerically, but comparing the successful transmission rate of any policy against this naive policy is important, as any efficient learning algorithm should outperform it.

@@

We investigate here the optimal policy that can be achieved if the dynamic devices have a perfect knowledge of everything, and a fully centralized decision making\footnote{~This optimal policy needs an \emph{oracle} seeing the entire system, and affecting all the dynamic devices, once and for all, in order to avoid any signaling overhead.} is possible.
We want to find the stationary repartition of devices into channels that maximizes the probability of having a successful transmission.

If the oracle draws once uniformly at random a configuration of dynamic devices, with $D_i$ devices affected to channel $i$ is fixed (in time, \ie, stationary),
then this probability is computed as before:
\begin{align}\label{eq:41:prob_col}
    \Pr(\text{success}|\text{sent})
    & = \sum_{i=1}^{K} \Pr(\text{success}|\text{sent in channel}\;i) \; \Pr(\text{sent in channel}\;i) \nonumber \\
    & = \sum_{i=1}^{K} \underbrace{(1 - p)^{D_i - 1}}_{\;\;D_i - 1 \;\text{others}\;\;} \times \underbrace{(1 - p)^{S_i}}_{\;\;\text{No static device}\;\;} \times \underbrace{ D_i / D }_{\;\;\text{Sent in channel}\; i\;\;}.
\end{align}

Consequently, the optimal allocation vector $(D_1,\dots,D_{K})$ is the solution of the following real-valued constraint optimization problem:
%
\begin{subequations}\label{eq:41:prob}
\begin{align}
    % \begin{cases}
    \underset{D_1,\dots,D_{K}}{\arg\max}\; & \sum_{i=1}^{K} D_i (1 - p)^{S_i + D_i -1}, \label{eq:41:optPb}\\
    \text{such that}\;\; & \sum_{i=1}^{K} D_i = D, \label{eq:41:eqCstr}\\
    & D_i \geq 0 \qquad \forall i\in\llbracket 1;K\rrbracket . \label{eq:41:ineqCstr}
    % \end{cases}
\end{align}
\end{subequations}

@@

    It gives a closed form expression for the optimal solution $D_i^*(\lambda)$, depending on the system parameters, and the unknown Lagrange multiplier $\lambda \in \mathbb{R}$.

    \begin{equation}\label{eq:41:Dilambda}
        D_i^*(\lambda) = \left(\frac{1}{\log(1-p)}\left[ \mathcal{W}\left(\frac{\lambda e}{(1-p)^{S_i-1}} \right)-1 \right]\right)^{\dag} .
    \end{equation}
\end{proposition}

@@

Where $(a)^{\dag} = \max(a,0)$, and $\mathcal{W}$ denotes the $\mathcal{W}$-Lambert function which is the reciprocal bijection of $x \mapsto xe^x$ on $\mathbb{R^+} = [0, +\infty)$ \cite{Corless96}.
Moreover, condition \eqref{eq:41:eqCstr} implies that the Lagrange multiplier $\lambda$ is the solution of the constraint
\begin{equation}\label{eq:41:constraintLambda}
 \sum_{i=1}^{K} D_i^*(\lambda) = D.
\end{equation}

@@

While the stochastic MAB model has been used to describe some aspects of Cognitive Radio systems, it is in principle not suitable for our IoT model, due to the non-stationarity of the channels occupancy caused by the learning policy used by dynamic objects.

In our model, every dynamic device implements its own learning algorithm, \emph{independently}.
For one device, the time $t$ is the number of time it accessed the network (following its Bernoulli transmission process, \ie, its duty cycle), \emph{not} the total number of time slots from the beginning, as rewards are only obtained after a transmission, and IoT objects only transmit sporadically, due to low transmission duty cycles.

@@

Considered alone, each dynamic device implements a learning algorithm to play a bandit game, the device is consequently a dynamic device. In each time slot, if it has to communicate (which happens with probability $p$), then it chooses a channel and it receives a reward $1$ if the transmission is successful, $0$ otherwise.
Each device aims at maximizing the sum of the rewards collected during its communication instants, which shall indeed maximize the fraction of successful transmissions. Besides the modified time scale (rewards are no longer collected at every time step), this looks like a bandit problem.
However, it cannot be modeled as a stochastic MAB, as the rewards are clearly \emph{not} \iid: they not only depend on the (stationary, \iid) behavior of the static devices, but also on the behavior of other dynamic devices, that is not stationary (because of learning).

@@

The \UCB{} algorithm uses a parameter $\alpha > 0$, originally $\alpha$ was set to $2$ \cite{Auer02}, but empirically $\alpha = 1/2$ is known to work better (uniformly across problems), and $\alpha > 1/2$ is advised by the theory \cite{Bubeck12}.
%
In our model, every dynamic dynamic device implements its own \UCB{} algorithm, \emph{independently}.
For one dynamic device, the time $t$ is the total number of sent messages from the beginning, as rewards are only obtained after a transmission.

The TS algorithm, although being simple and easy to implement, is known to perform well for stochastic problems, for which it was proven to be asymptotically optimal \cite{AgrawalGoyal11,Kaufmann12Thompson}.
It is known to be empirically efficient, and for these reasons it has been used successfully in various applications, including on problems from Cognitive Radio \cite{Toldov16,Mitton16}, and also in previous work on decentralized IoT-like networks \cite{Darak16}.

@@

The main difference between the two models is the fact that in this Chapter, devices do not transmit their messages at every time step, but following a random activation process (with a fixed transmission probability $p$).
Active devices denote static and dynamic devices that have to transmit at a fixed time,
and their random activation patter makes the number of active devices a random variable.
Analyzing multi-player MAB models under this hypothesis is much harder, and is left as a future work.

>>>>

As explained before, setting up the future Internet of Things (IoT) networks will require to support more and more communicating devices.
In this Section, we prove that intelligent devices in unlicensed bands can use Multi-Armed Bandit (MAB) learning algorithms to improve resource exploitation.
%
We evaluate the performance of two classical MAB learning algorithms, \UCB{} and Thomson Sampling, to handle the decentralized decision-making of Spectrum Access, applied to IoT networks. We also evaluate the learning performance when the number of intelligent end-devices grows.
%
We show that using learning algorithms does help to fit more devices in such networks, even when all end-devices are intelligent and are dynamically changing channel.
In the studied scenario, stochastic MAB learning provides an up to $16\%$ gain in term of successful transmission probabilities, and has near optimal performance even in non-stationary settings with a majority of intelligent devices.

@@

The aim of this Section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases, and the stochastic hypothesis is more and more violated.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could have been generated by devices using other standards as well.
We first evaluate the probability of collision if dynamic devices randomly select channels (that is, a naive approach), and if a centralized controller optimally distribute them in the channels at the beginning of the scenario (that is, an ideal approach).
Then, these three reference scenarios allow to evaluate the performance of \UCB{} and TS algorithms in a decentralized network, in terms of successful communication rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic.

\paragraph{Outline}
%
The rest of this Section is organized as follows. The system model is introduced in Section~\ref{sub:41:systemModel}. The reference policies are described in \ref{sub:41:threeReferencePolicies}, and our proposed sequential learning policies are introduced in \ref{sub:41:sequentialPolicies}.
Numerical results are presented in \ref{sub:41:numericalResults}.

@@

We present our system model, which consists of one gateway and many IoT devices, using a frequency- and time- slotted protocol.

\paragraph{One gateway and many devices}
%
We consider the system model presented in Figure~\ref{fig:41:system_model1}, where a set of objects all sends uplink packets to the (unique) network gateway.
Messages can be sent for instance centered around the $433.5\;\mathrm{MHz}$ ISM bands (as it is chosen for the implementation presented in Section~\ref{sec:4:gnuradio}).
The communication between IoT devices and this gateway is done through a simple pure ALOHA-based protocol where devices transmit uplink packets of fixed duration whenever they want.
%
The devices can transmit their packets in $K\geq2$ channels (\eg, $K=4$). In the case where the gateway receives an uplink in one channel, it transmits an acknowledgement to the end-device in the same channel, after a fixed delay (of $1\;\mathrm{s}$ in our implementation).

These communications operate in unlicensed ISM bands and, consequently, they can suffer from interference generated by uncoordinated neighboring networks. This interfering traffic is uncontrolled, and is most likely unevenly distributed over the $K$ different channels.
%
To simulate networks designed for the Internet of Things (IoT), we consider a protocol \textbf{with no sensing}, no repetition of uplink messages, and where the gateway is in charge of sending back an acknowledgement, after some fixed-time delay (typically $1\;\mathrm{s}$), to any object who succeeded in sending successfully an uplink packet (but only to the objects being paired with this base station in this network).
%
By considering a small number of wireless channels (\eg, $K = 4$) and a PHY layer configuration (\ie, modulation, waveform, etc), and in case of a non-uniform traffic in the different channels,
the object can improve their usage of the network if they are able to \emph{learn} on the fly the best channels to use (\ie, the most vacant one).

@@

\paragraph{Slotted protocol}
%
For simplicity, and for consistency with the rest of this thesis, we only consider a discrete protocol (\ie, slotted).
As illustrated in Figure \ref{fig:41:protocol}, we suppose a slotted protocol, in both time and frequency.
All devices share a synchronized time, and know in advance $K$, the finite number of available RF channels.
In each time slot, the devices try to send packets to the unique Base Station, which listens continuously to all channels, following an ALOHA-based communication (\ie, no sensing).
Each time slot is divided in two parts: first for uplink communications in which data packets are sent by end-devices to the base station.
In one channel, if only one packet is sent in this part of the slot, the base station can decode it and sends an acknowledgement to the device in the second part (on the same channel).
If two or more devices send an uplink packet in the same slot, the uplink packets collide (\ie, there is a \emph{collision}), and the acknowledgement \Ack{} is not transmitted.
In other words, if the gateway cannot successfully decode the incoming (uplink) messages because they were corrupted by collisions, it does not send any \Ack{} back.
This way, no collision can occur on the downlink messages, easing the analysis of collisions.

\begin{figure}[!t]
    \centering
    \includegraphics[scale=0.40]{protocol.eps}
    \caption{The considered time-frequency slotted protocol. Each frame is composed by a fix-duration \textcolor{blue}{uplink slot} in which the end-devices transmit their \textcolor{blue}{(uplink) packets}. If a packet is well received, the base station replies by transmitting an \textcolor{darkgreen}{\Ack}, after the ack delay.}
    \label{fig:41:protocol}
\end{figure}

\paragraph{Static vs dynamic devices}
%
There are two types of end-devices in the network:
\begin{itemize}
    \item
    \emph{Static} end-devices have poor RF abilities, and each of them uses only one channel to communicate with the base station. Their choice is assumed to be fixed in time (\ie, stationary). The traffic generated by these devices is considered as an interfering traffic for other devices.
    \item
    \emph{Dynamic} (or \emph{smart}) end-devices have richer RF abilities, they can use all the available channels, by quickly re-configuring their RF transceiver on the fly. They can also store communication successes or failures they experienced in each channel, in order to change channel, possibly at every time slot.
    We also assume that the dynamic end-devices can run simple embedded decision making algorithms, and have limited but reasonable computing as well as storage capacities.
\end{itemize}

There are $K \geq 2$ channels, $D \geq 0$ dynamic end-devices, and $S \geq 0$ static devices.
Furthermore, in channel $i \in \llbracket 1; K \rrbracket$ there are $0 \leq S_i \leq S$ static devices (so $S = \sum_{i=1}^{K} S_i$).

@@

\paragraph{Dense networks and an hypothesis}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the base station, we assume devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for IoT networks\footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
Indeed, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ objects, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall i, S_i > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero than using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$ to have an average of active objects not larger than the number of channels.

\paragraph{MAB model}
%
We consider the network from the point of view of one dynamic device.
Following its Bernoulli emission process, every times a dynamic device has to communicate with the gateway,
it has to choose one channel (at each transmission $t \geq 1, t \in \mathbb{N}$), denoted as $C(t) = k \in\{1,\dots,K\}$.
Then, the dynamic device starts waiting in this channel $C(t)$ for an acknowledgement sent by the gateway.
Before sending another message (\ie, at time $t+1$), the dynamic device knows if it received or not this \Ack{} message.
%
For this reason, selecting channel (or arm) $k$ at time $t$ yields a (random) feedback, called a (binary) \emph{reward}, $r_k(t) \in \{0,1\}$, being $0$ if no \Ack{} was received before the next message, or $1$ if \Ack{} was successfully received.
The goal of the dynamic device is to minimize its packet loss ratio, or equivalently, it is to maximize its number of successful transmissions, or its cumulative reward,
$r_{1 \dots T} := \sum_{t = 1}^T r_{C(t)}(t),$
as it is usually done in MAB problems.

This problem is a special case of the stochastic MAB (see Chapter~\ref{chapter:2}).
% , where the sequence of rewards drawn from a given arm $k$ is assumed to be  \emph{i.i.d.}, under some distribution $\nu_k$, that has a mean $\mu_k$.
% Several types of reward distributions have been considered in the literature, and rewards are binary in our model.
We consider only Bernoulli distributions, in which $r_k(t) \sim \mathrm{Bern}(\mu_k)$, that is, $r_k(t) \in \{0,1\}$ and $\mathbb{P}(r_k(t) = 1) = \mu_k \in [0,1]$.
Contrary to many previous work done in the CR field (\eg, Opportunistic Spectrum Access),
the reward $r_k(t)$ does \emph{not} come from a sensing phase before sending the $t$-th message, as it would do for any ``listen-before-talk'' model.
In our model, rewards rather come from receiving an acknowledgement from the gateway, between the $t$-th and $t+1$-th messages.

The problem parameters are $\mu_1,\dots,\mu_K$, the mean availability of the channels, and they are of course unknown to each dynamic device, so to maximize its cumulated reward, it must learn the distributions of the channels, in order to be able to progressively improve its successful communication rate.
%
This requires to tackle the so-called \emph{exploration-exploitation dilemma}: a device has to try all arms a sufficient number of times to get a robust estimate of their qualities, while not selecting the worst arms too many times.
%
The goal is thus to design a new or chose an existing simple sequential algorithm, to be applied identically and independently by each dynamic device, in a fully distributed setting (each device runs its own algorithm, from its observations), in order to minimize collisions and maximize the fraction of successful transmissions of all the dynamic devices.

Before modeling this goal with a \emph{multi-armed bandit problem}, we present some natural baseline (reference) policies.

@@

We present three different policies that can be used to assess the efficiency of the learning algorithms presented later on.
The first one is naive but can be used in practice, while the two others are very efficient but require full knowledge on the system (\ie, an oracle) and are thus unpractical.
%
They are however useful for our numerical simulations, to show that our MAB-based approaches quickly learn to perform almost optimally.

@@

We derive here the probability of having a successful transmission, for a dynamic device, in the case where all the dynamic devices make a purely random channel selection (\ie, uniform on $i \in \llbracket 1; K \rrbracket$).
This reflects a naive policy that could be implemented by all the dynamic devices, and it provides a reference scenario to compare against.

In this case, for one dynamic device, a successful transmission happens if it is the only device to choose channel $i$, at that time slot.
The probability of successful transmission is computed as follows, because the $S_i$ static devices in each channel $i$ are assumed to be independent, and static and dynamic devices are assumed to \emph{not} transmit at each time $t$ with a fixed probability $1-p$:
\begin{equation}
    \Pr(\text{success}|\text{sent}) = \sum_{i=1}^{K} \underbrace{\Pr(\text{success}|\text{sent in channel}\;i)}_{\text{No one else sent in channel}\; i} \; \underbrace{\Pr(\text{sent in channel}\,i)}_{= 1/K, \text{ by uniform choice}}
\end{equation}

All dynamic devices follow the same policy in this case, so the probability of transmitting at that time in channel $i$ for any dynamic device is $p / K$, and there are $D-1$ other dynamic devices.
As they are independent, the probability that no other dynamic device sent in $i$
% $\Pr(\text{no other dynamic device sent in}\;i)$,
is $q = \Pr(\bigcap_{k=1}^{D-1} \text{device}\;k\;\text{did not sent in}\;i) = \prod_{k=1}^{D-1} \Pr(\text{device}\;k\;\text{did not sent in}\;i)$. And $\Pr(\text{device}\;k\;\text{sent in}\;i) = p \times 1 / K$, by uniform choice on channels and the Bernoulli emission hypothesis. So $q = \prod_{k=1}^{D-1} (1 - p/K) = (1-p/K)^{D-1}$. Thus we can conclude,
%
\begin{align}\label{eq:41:strategynaive}
    \Pr(\text{success}|\text{sent})
    & = \sum_{i=1}^{K} \underbrace{(1 - p / K)^{D-1}}_{\text{No other dynamic device}} \times \underbrace{(1-p)^{S_i}}_{\text{No static device}} \times\; \frac{1}{K} \nonumber \\
    & = \frac{1}{K} \left(1-\frac{p}{K}\right)^{D-1} \sum_{i=1}^{K} (1-p)^{S_i} .
\end{align}
This expression \eqref{eq:41:strategynaive} is constant (in time), and easy to compute numerically, but comparing the successful transmission rate of any policy against this naive policy is important, as any efficient learning algorithm should outperform it
(maybe after a long enough initial learning period).

@@

We investigate here the optimal policy that can be achieved if the dynamic devices have a perfect knowledge of the repartition of static devices (\ie, $(S_i)_i$ or equivalently, $(\mu_i)_i$), and a fully centralized decision making\footnote{~This optimal policy needs an \emph{oracle} seeing the entire system, and affecting all the dynamic devices, once and for all, for a fixed stationary scenario, in order to avoid any signaling overhead.} is possible.
We want to find the stationary repartition of devices into channels that maximizes the probability of having a successful transmission.

If the oracle chooses a fixed configuration of dynamic devices, with $D_i$ devices affected to channel $i$ is fixed (in time, \ie, stationary),
then this probability is computed as before:
\begin{align}\label{eq:41:prob_col}
    \Pr(\text{success}|\text{sent})
    & = \sum_{i=1}^{K} \Pr(\text{success}|\text{sent in channel}\;i) \; \Pr(\text{sent in channel}\;i) \nonumber \\
    & = \sum_{i=1}^{K} \underbrace{(1 - p)^{D_i - 1}}_{\;\;D_i - 1 \;\text{others}\;\;} \times \underbrace{(1 - p)^{S_i}}_{\;\;\text{No static device}\;\;} \times \underbrace{ D_i / D }_{\;\;\text{Sent in channel}\; i\;\;}.
\end{align}

Consequently, an optimal allocation vector $(D_1,\dots,D_{K})$ is a solution of the following real-valued constraint optimization problem:
%
\begin{subequations}\label{eq:41:prob}
\begin{align}
    % \begin{cases}
    \underset{D_1,\dots,D_{K}}{\arg\max}\; & \sum_{i=1}^{K} D_i (1 - p)^{S_i + D_i -1}, \label{eq:41:optPb}\\
    \text{such that}\;\; & \sum_{i=1}^{K} D_i = D, \label{eq:41:eqCstr}\\
    & D_i \geq 0 \qquad \forall i\in\llbracket 1;K\rrbracket . \label{eq:41:ineqCstr}
    % \end{cases}
\end{align}
\end{subequations}

@@

    It gives a closed form expression for the (unique) optimal solution $D_i^*(\lambda)$, depending on the system parameters, and the unknown Lagrange multiplier $\lambda \in \mathbb{R}$.

    \begin{equation}\label{eq:41:Dilambda}
        D_i^*(\lambda) = \left(\frac{1}{\log(1-p)}\left[ \mathcal{W}\left(\frac{\lambda e}{(1-p)^{S_i-1}} \right)-1 \right]\right)^{+} .
    \end{equation}
\end{proposition}

@@

Where $(a)^{+} = \max(a,0)$, and $\mathcal{W}$ denotes the $\mathcal{W}$-Lambert function which is the reciprocal bijection of $x \mapsto x e^x$ on $\mathbb{R^+} = [0, +\infty)$ \cite{Corless96}.
Moreover, condition \eqref{eq:41:eqCstr} implies that the Lagrange multiplier $\lambda$ is the solution of the constraint
\begin{equation}\label{eq:41:constraintLambda}
 \sum_{i=1}^{K} D_i^*(\lambda) = D.
\end{equation}

@@

While the stochastic MAB model has been used to describe some aspects of Cognitive Radio systems, it is in principle not suitable for our IoT model, due to the non-stationarity of the channels occupancy caused partly by the learning policy used by dynamic objects, but mainly by their random activation processes.

In our model, every dynamic device implements its own learning algorithm, \emph{independently}.
For one device, the time $t$ refers to the number of time it accessed the network (following its Bernoulli transmission process, \ie, its duty cycle), \emph{not} the total number of time slots from the beginning, as rewards are only obtained after a transmission, and IoT objects only transmit sporadically, due to low transmission duty cycles.

@@

Considered alone, each dynamic device implements a learning algorithm to play a bandit game.
In each time slot, if it has to communicate (which happens with probability $p$), then it chooses a channel and it receives a reward $1$ if the transmission is successful, $0$ otherwise.
Each device aims at maximizing the sum of the rewards collected during its communication instants, which shall indeed maximize the fraction of successful transmissions. Besides the modified time scale (rewards are no longer collected at every time step), this looks like a usual bandit problem.
However, it cannot be modeled as a stochastic MAB, as the rewards are clearly \emph{not} \iid: they not only depend on the (stationary, \iid) behavior of the static devices, but also on the behavior of other dynamic devices, that is not stationary (because of learning and random activation of each devices).

@@

The two algorithms, \UCB{} and Thompson sampling (TS), were presented in Section~\ref{sec:2:famousMABalgorithms}.
The \UCB{} algorithm uses a parameter $\alpha > 0$, originally $\alpha$ was set to $2$ \cite{Auer02}, but empirically $\alpha = 1/2$ is known to work better (uniformly across problems), and $\alpha > 1/2$ is advised by the theory \cite{Bubeck12}.
%
In our model, every dynamic dynamic device implements its own \UCB{} algorithm, \emph{independently}.
For one dynamic device, the time $t$ is the total number of sent messages from the beginning, as rewards are only obtained after a transmission.
%
The TS algorithm, although being simple and easy to implement, is known to perform well for stochastic problems, for which it was proven to be asymptotically optimal \cite{AgrawalGoyal11,Kaufmann12Thompson}.
It is known to be empirically efficient, and for these reasons it has been used successfully in various applications, including on problems from Cognitive Radio \cite{Toldov16,Mitton16}, and also in previous work on decentralized IoT-like networks \cite{Darak16}.

@@

The main difference between the two models is the fact that in this Chapter, devices do not transmit their messages at every time step, but following a random activation process (with a fixed transmission probability $p < 1$).
Active devices denote static and dynamic devices that have to transmit at a fixed time,
and their random activation patter makes the number of active devices a random variable.
Analyzing multi-player MAB models under this hypothesis is much harder, and is left as a future work.
